\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-19 paper.qmd
Added intial analysis for PBMC

```{r}
hex_param_vis <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
    geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
    geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "#31a354") +
    geom_point(data = all_centroids_df |> dplyr::filter(hexID == 1), aes(x = c_x, y = c_y), color = "black") + ## starting point
    geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = r2), fill = "white", color = "black", alpha = 0, linewidth = 1) +
    geom_segment(x = 0, y = -0.35, xend = 1, yend = -0.35,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+
    annotate("text", x=0.5, y=-0.45, label= "r1", color = "black") +
    geom_segment(x = -0.25, y = 0, xend = -0.25, yend = r2,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ 
    annotate("text", x=-0.35, y=1, label= "r2", color = "black") +
    geom_segment(x = -0.1 + 0.2087578, y = -0.15, xend = -0.1 + 0.2087578*2, yend = -0.15,
                   arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a1 = 0.2087578
    annotate("text", x=0.2, y=-0.25, label= "a1", color = "black") +
    geom_segment(x = -0.15, y = -0.1*r2 + 0.1807896*2, xend = -0.15, yend = -0.1*r2 + 0.1807896*3,
                     arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a2 = 0.1807896
    annotate("text", x=-0.15, y=0.45, label= "a2", color = "black") +
    annotate("text", x=-0.15, y=-0.35, label= "(s1, s2)", color = "black") ## starting point
```

```{r}
#| label: fig-hex-param
#| fig-cap: "Notations for hexagonal grid configurations."
#| fig-width: 3
#| fig-height: 3
#| fig-pos: H

@@

### pbmc

- NLDR view used to illustrate clusters
- Use our method to assess is it a reasonable representation
- Demonstrate that it is not
- Illustrate how to use our method to get a better representation

@@

<!-- UMAP parameters suggested by the author(n_neighbors = 30, min_dist = 0.3)-->
```{r}
#| fig-cap: "2D layout from UMAP applied for the PBMC3k dataset. Is this a best representation of the original data?"
#| label: fig-umapauthor
#| fig-width: 3
#| fig-height: 3
#| fig-pos: H

@@

@@

<!--bin1 = 30, bin2 = 29, b = 870, non_empty = 132-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in the 2D space overlaid on UMAP data, and (b) high-D model error in model space."
#| label: fig-model-pbmc-author
#| fig-pos: H

@@

@@

error_plot_pbmc <- ggplot(error_df, 
                              aes(x = b_non_empty, 
                                  y = log(error), 
                                  group = type, 
                                  colour = type)) + 
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = 137, linetype="solid", 
             color = "black", size=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#984ea3',
                              '#ff7f00','#a6cee3','#a65628','#f781bf')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")

@@

```{r}
#| fig-cap: "Absolute error from UMAP and tSNE applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to 137, the lowest error occurred with the parameters perplexity: 30."
#| label: fig-abserrorPBMC
#| fig-pos: H
#| out-width: 100%

@@

@@

@@

<!--bin1 = 15, bin2 = 20, b = 300, non_empty = 137-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in the 2D space overlaid on tSNE data, and (b) high-D model error in model space."
#| label: fig-model-pbmc
#| fig-pos: H

>>>>

```{r}
hex_param_vis <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
    geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
    geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "#31a354") +
    geom_point(data = all_centroids_df |> dplyr::filter(hexID == 1), aes(x = c_x, y = c_y), color = "black") + ## starting point
    geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = r2), fill = "white", color = "black", alpha = 0, linewidth = 1) +
    geom_segment(x = 0, y = -0.35, xend = 1, yend = -0.35,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+
    annotate("text", x=0.5, y=-0.45, label= "r1", color = "black") +
    geom_segment(x = -0.25, y = 0, xend = -0.25, yend = r2,
                 arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ 
    annotate("text", x=-0.35, y=1, label= "r2", color = "black") +
    geom_segment(x = -0.1 + 0.2087578, y = -0.15, xend = -0.1 + 0.2087578*2, yend = -0.15,
                   arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a1 = 0.2087578
    annotate("text", x=0.2, y=-0.25, label= "a1", color = "black") +
    geom_segment(x = -0.15, y = -0.1*r2 + 0.1807896*2, xend = -0.15, yend = -0.1*r2 + 0.1807896*3,
                     arrow = arrow(length = unit(0.03, "npc"), ends = "both"), color = "black")+ # a2 = 0.1807896
    annotate("text", x=-0.15, y=0.45, label= "a2", color = "black") +
    annotate("text", x=-0.15, y=-0.35, label= "(s1, s2)", color = "black") + ## starting point
  coord_fixed()
```

```{r}
#| label: fig-hex-param
#| fig-cap: "Notations for hexagonal grid configurations."
#| out-height: 30%
#| fig-pos: H

@@

### pbmc
<!--
- NLDR view used to illustrate clusters
- Use our method to assess is it a reasonable representation
- Demonstrate that it is not
- Illustrate how to use our method to get a better representation
-->

In the field of single-cell studies, a common analytical task involves clustering to identify groups of cells with similar expression profiles. Analysts often turn to NLDR techniques to verify and identify these clusters and explore developmental trajectories. To illustrate the importance of NLDR techniques and parameter selection in identifying clusters, Human Peripheral Blood Mononclear Cells (PBMC3k) dataset [@chen2023] is used. In a study by @chen2023, this dataset was used to demonstrate how UMAP represents clusters (see @fig-umap-author). As shown in @fig-umap-author, there are three distinct and well-separated clusters.

@@

<!-- UMAP parameters suggested by the author (n_neighbors = 30, min_dist = 0.3)-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from UMAP applied for the PBMC3k dataset. Is this a best representation of the original data? The parameter setting is n_neighbors = 30, min_dist = 0.3."
#| label: fig-umap-author
#| out-height: 30%
#| fig-pos: H

@@

To determine whether the UMAP representation with the parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the \pD{} data. The figures in @fig-pbmc1_sc show three well-separated clusters, indicating that the suggested UMAP representation preserves the global structure (see @fig-umap-author). However, as shown in @fig-pbmc1_sc, these clusters are close to each other in \pD{}. Also, non-linear continuity patterns and high-density patches within the clusters are observed (see @fig-pbmc1_sc). Therefore, the suggested UMAP representation (see @fig-umap-author) does not accurately preserve the local structure of the PBMC3k dataset.

@@

<!--bin1 = 30, bin2 = 29, b = 870, non_empty = 132-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with UMAP, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distinct clusters. The $p\\text{-}D$ model errors are distributed along clusters."
#| label: fig-model-pbmc-author
#| fig-pos: H

@@

::: {#fig-pbmc1_sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_1.png){#fig-pbmc1_sc1}

![](figures/pbmc3k/sc_2.png){#fig-pbmc1_sc2}

![](figures/pbmc3k/sc_3.png){#fig-pbmc1_sc3}

Screen shots of the **langevitour** of the PBMC3k data set, shows the model-in-data space, a video of the tour animation is available at (<https://youtu.be/Gnhh4hfsbbg>).
:::

In order to find a reasonable NLDR representation for the PBMC3k dataset, we calculated the absolute error for different numbers of non-empty bins using various NLDR techniques and different parameter settings (see @fig-pbmc-abserror). After analyzing the results, we found that tSNE with a perplexity set to $30$ had the lowest error when the number of non-empty bins was $137$. Therefore, tSNE with a perplexity of $30$, which is the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset.

@@

error_plot_pbmc <- ggplot(error_df, 
                              aes(x = b_non_empty, 
                                  y = log(error), 
                                  group = type, 
                                  colour = type)) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 137, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#984ea3',
                              '#ff7f00','#a6cee3','#a65628','#f781bf')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins") +
  theme(axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90))

@@

```{r}
#| fig-cap: "Absolute error from UMAP and tSNE applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to $137$, the lowest error occurred with the parameters perplexity: $30$."
#| label: fig-pbmc-abserror
#| fig-pos: H
#| out-width: 50%

@@

As shown in @fig-tsne-suggest, there are three well-separated clusters, although they are located close to each other. Additionally, non-linear structures can also be observed within the clusters (see @fig-model-pbmc (a)). In this manner, tSNE was able to capture the data structure for the PBMC3k dataset that UMAP failed to do.

@@

<!--tSNE with perplexity:30-->
```{r}
#| echo: false
#| fig-cap: "$2\\text{-}D$ layout from tSNE applied for the PBMC3k dataset.  Is this a best representation of the original data? The parameter setting is perplexity=30."
#| label: fig-tsne-suggest
#| fig-pos: H
#| out-width: 40%

tsne_pbmc 
```

We then fit the model for tSNE, and visualize the resultant model in the \pD{} data space. The model shows a quirk, as shown in @fig-pbmc2_sc. All three clusters are connected by an edge except the small and large clusters. Because the clusters are so close in \gD{}, they attempt to maintain the structure in \pD{} as well. This is evident that tSNE with perplexity $30$ provides a reasonable representation of PBMC3k data.

@@

<!--bin1 = 15, bin2 = 20, b = 300, non_empty = 137-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with tSNE, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distinct clusters. The $p\\text{-}D$ model errors are distributed along clusters, but most low $p\\text{-}D$ model errors present in the large cluster."
#| label: fig-model-pbmc
#| fig-pos: H
#| out-height: 30%

### 2024-05-19 paper.tex
Added intial analysis for PBMC

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-hex-param-1.pdf}

@@

\begin{itemize}
\tightlist
\item
  NLDR view used to illustrate clusters
\item
  Use our method to assess is it a reasonable representation
\item
  Demonstrate that it is not
\item
  Illustrate how to use our method to get a better representation
\end{itemize}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-umapauthor-1.pdf}

@@

\caption{\label{fig-umapauthor}2D layout from UMAP applied for the
PBMC3k dataset. Is this a best representation of the original data?}

@@

@@

\caption{\label{fig-model-pbmc-author}(a) Model generated in the 2D
space overlaid on UMAP data, and (b) high-D model error in model space.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-abserrorPBMC-1.pdf}

@@

\caption{\label{fig-abserrorPBMC}Absolute error from UMAP and tSNE
applied to training PBMC3k dataset with diffferent parameter choices.
What is the best parameter choice to create the model? The residual plot
have a steep slope at the beginning, indicating that a smaller number of
non-empty bins causes a larger amount of error. Then, the slope
gradually declines or level off, indicating that a higher number of
non-empty bins generates a smaller error. Using the elbow method, it was
observed that when the number of non-empty bins is set to 137, the
lowest error occurred with the parameters perplexity: 30.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-model-pbmc-1.pdf}

@@

\caption{\label{fig-model-pbmc}(a) Model generated in the 2D space
overlaid on tSNE data, and (b) high-D model error in model space.}

>>>>

\includegraphics[width=1\textwidth,height=0.3\textheight]{paper_files/figure-pdf/fig-hex-param-1.pdf}

@@

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
Analysts often turn to NLDR techniques to verify and identify these
clusters and explore developmental trajectories. To illustrate the
importance of NLDR techniques and parameter selection in identifying
clusters, Human Peripheral Blood Mononclear Cells (PBMC3k) dataset
\citep{chen2023} is used. In a study by \citet{chen2023}, this dataset
was used to demonstrate how UMAP represents clusters (see
Figure~\ref{fig-umap-author}). As shown in Figure~\ref{fig-umap-author},
there are three distinct and well-separated clusters.

@@

\includegraphics[width=1\textwidth,height=0.3\textheight]{paper_files/figure-pdf/fig-umap-author-1.pdf}

@@

\caption{\label{fig-umap-author}\(2\text{-}D\) layout from UMAP applied
for the PBMC3k dataset. Is this a best representation of the original
data? The parameter setting is n\_neighbors = 30, min\_dist = 0.3.}

@@

To determine whether the UMAP representation with the parameter choice
suggested by \citet{chen2023} preserves the original data structure, we
visualize the model constructed with UMAP overlaid on the \pD{} data.
The figures in Figure~\ref{fig-pbmc1_sc} show three well-separated
clusters, indicating that the suggested UMAP representation preserves
the global structure (see Figure~\ref{fig-umap-author}). However, as
shown in Figure~\ref{fig-pbmc1_sc}, these clusters are close to each
other in \pD{}. Also, non-linear continuity patterns and high-density
patches within the clusters are observed (see
Figure~\ref{fig-pbmc1_sc}). Therefore, the suggested UMAP representation
(see Figure~\ref{fig-umap-author}) does not accurately preserve the
local structure of the PBMC3k dataset.

@@

\caption{\label{fig-model-pbmc-author}(a) Model generated in
\(2\text{-}D\) with UMAP, and (b) \(p\text{-}D\) model error in
\(2\text{-}D\). The \(2\text{-}D\) model shows three well-separated
distinct clusters. The \(p\text{-}D\) model errors are distributed along
clusters.}

@@

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_1.png}

@@

\subcaption{\label{fig-pbmc1_sc1}}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_2.png}

\subcaption{\label{fig-pbmc1_sc2}}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_3.png}

\subcaption{\label{fig-pbmc1_sc3}}

\caption{\label{fig-pbmc1_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model-in-data space, a video of the tour
animation is available at (\url{https://youtu.be/Gnhh4hfsbbg}).}

In order to find a reasonable NLDR representation for the PBMC3k
dataset, we calculated the absolute error for different numbers of
non-empty bins using various NLDR techniques and different parameter
settings (see Figure~\ref{fig-pbmc-abserror}). After analyzing the
results, we found that tSNE with a perplexity set to \(30\) had the
lowest error when the number of non-empty bins was \(137\). Therefore,
tSNE with a perplexity of \(30\), which is the default parameter
setting, is considered as a reasonable representation for the PBMC3k
dataset.

\includegraphics[width=0.5\textwidth,height=\textheight]{paper_files/figure-pdf/fig-pbmc-abserror-1.pdf}

\caption{\label{fig-pbmc-abserror}Absolute error from UMAP and tSNE
applied to training PBMC3k dataset with diffferent parameter choices.
What is the best parameter choice to create the model? The residual plot
have a steep slope at the beginning, indicating that a smaller number of
non-empty bins causes a larger amount of error. Then, the slope
gradually declines or level off, indicating that a higher number of
non-empty bins generates a smaller error. Using the elbow method, it was
observed that when the number of non-empty bins is set to \(137\), the
lowest error occurred with the parameters perplexity: \(30\).}

As shown in Figure~\ref{fig-tsne-suggest}, there are three
well-separated clusters, although they are located close to each other.
Additionally, non-linear structures can also be observed within the
clusters (see Figure~\ref{fig-model-pbmc} (a)). In this manner, tSNE was
able to capture the data structure for the PBMC3k dataset that UMAP
failed to do.

\includegraphics[width=0.4\textwidth,height=\textheight]{paper_files/figure-pdf/fig-tsne-suggest-1.pdf}

\caption{\label{fig-tsne-suggest}\(2\text{-}D\) layout from tSNE applied
for the PBMC3k dataset. Is this a best representation of the original
data? The parameter setting is perplexity=30.}

We then fit the model for tSNE, and visualize the resultant model in the
\pD{} data space. The model shows a quirk, as shown in
Figure~\ref{fig-pbmc2_sc}. All three clusters are connected by an edge
except the small and large clusters. Because the clusters are so close
in \gD{}, they attempt to maintain the structure in \pD{} as well. This
is evident that tSNE with perplexity \(30\) provides a reasonable
representation of PBMC3k data.

\includegraphics[width=1\textwidth,height=0.3\textheight]{paper_files/figure-pdf/fig-model-pbmc-1.pdf}

\caption{\label{fig-model-pbmc}(a) Model generated in \(2\text{-}D\)
with tSNE, and (b) \(p\text{-}D\) model error in \(2\text{-}D\). The
\(2\text{-}D\) model shows three well-separated distinct clusters. The
\(p\text{-}D\) model errors are distributed along clusters, but most low
\(p\text{-}D\) model errors present in the large cluster.}

@@

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_4.png}

@@

\subcaption{\label{fig-pbmc2_sc4}}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_5.png}

\subcaption{\label{fig-pbmc2_sc5}}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_6.png}

\subcaption{\label{fig-pbmc2_sc6}}

\caption{\label{fig-pbmc2_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/R0EroJcGbas}).}
